---
title: "Introduction to sdmhelpers"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to sdmhelpers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(sdmhelpers)
library(sf)
library(ggplot2)
```

# Overview

`sdmhelpers` provides helper functions for species distribution modelling (SDM)
workflows, with particular emphasis on presence-background models implemented
with `maxnet`. The package contains tools for constructing sampling-bias
surfaces from target-group records, weighting target-group observations by
seasonal or combined seasonal and diel overlap, selecting complete spatial
folds for independent testing, screening environmental predictors, and
evaluating predictions using the true skill statistic (TSS).

Presence-only SDMs are sensitive to differences between the sampling process
that generated species observations and the process used to select background
locations. A commonly used approach is therefore to select or weight background
data so that they reflect similar sampling bias to the presence data
[@phillips2009; @elith2011]. `sdmhelpers` extends this idea by allowing records
from a target group to contribute unequally according to their temporal
similarity to the focal species.

The examples below use datasets distributed with `sdmhelpers`. They are reduced
or pseudonymised datasets intended for demonstrating package functionality, not
for biological inference.

`sdmhelpers` was developed as part of the project *HiQBioDiv:
High-resolution quantification of biodiversity for conservation and
management*, funded by the Latvian Council of Science
(Ref. No. VPP-VARAM-DABA-2024/1-0002).

The development version can be installed from GitHub with:

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("aavotins/sdmhelpers")
```

# Example data

Several example datasets are included in the package. The examples in this
vignette use bird, amphibian and bryophyte observations, spatial presence and background
locations, spatial blocks, a 100-m grid, known habitat-inventory dates, and an
example environmental raster.

```{r data}
data("aves_small", package = "sdmhelpers")
data("bryophyta_one", package = "sdmhelpers")
data("habitat_inventories", package = "sdmhelpers")
data("points100", package = "sdmhelpers")
data("example_grid", package = "sdmhelpers")
example_grid=terra::unwrap(example_grid)
data("example_presences", package = "sdmhelpers")
data("example_background", package = "sdmhelpers")
data("example_independent_blocks", package = "sdmhelpers")
data("egv_names", package = "sdmhelpers")
```

The environmental raster used later in the vignette is stored in
`inst/extdata`.

```{r env-path}
egv_file <- system.file("extdata",
                        "example_egv.tif",
                        package = "sdmhelpers")

egv_file
```

# Simple target-group bias surface

Target-group background sampling uses observations of taxa that are expected to
have been collected through a sampling process similar to that of the focal
species. If sampling effort is spatially biased, drawing background data from
the same biased sampling domain can reduce the contrast between presence and
background sampling processes [@phillips2009].

Here, one species from `aves_small` is treated as the focal species and all
remaining species form the target group. The focal species is chosen
programmatically as the species with the largest number of records so that the
example does not depend on a particular species name.

```{r simple-target-group}
target_species <- names(
  sort(table(aves_small$species), decreasing = TRUE)
)[1]

target_species

target_group <- aves_small
```

Multiple observations can occur in the same 100-m grid cell. For the spatial
bias surface, records are therefore counted by grid cell and supplied to
`kde_surface()` as weights.

```{r simple-target-group-points}
tg_counts <- aggregate(
  list(n_records = rep(1, nrow(target_group))),
  by = list(id100 = target_group$id100),
  FUN = sum
)

idx <- match(tg_counts$id100, points100$id)
keep <- !is.na(idx)

tg_points <- points100[idx[keep], ]
tg_points$n_records <- tg_counts$n_records[keep]

tg_points

```

`kde_surface()` first bins point weights to the reference raster and then
applies Gaussian smoothing. This makes the calculation practical for large
rasters and large bandwidths. Kernel density estimation in general is described
by @silverman1986.

The example grid uses EPSG:3059, whose map units are metres, so `sigma = 3000`
corresponds to a Gaussian bandwidth of 3 km. With `normalize = "pdf"`, the
resulting surface is normalized to integrate to one and represents the spatial
probability density of target-group sampling effort. It can therefore be used
as a relative sampling-bias surface for background sampling.


```{r simple-kde, eval=TRUE, fig.cap = "Unweighted target-group sampling-bias surface."}
simple_bias <- kde_surface(
  x = tg_points,
  weight_field = "n_records",
  sigma = 3000,
  ref = example_grid,
  normalize = "pdf",
  mask = TRUE
)

terra::plot(simple_bias, main = "Simple target-group bias")
```

This simple approach assumes that every target-group record is equally
informative about the sampling process affecting the focal species. The next
sections relax that assumption.


# Phenologically weighted target group

Species observed by the same people or recording scheme may nevertheless have
different seasonal detectability. A species recorded mainly in winter, for
example, may provide little information about sampling effort relevant to a
species recorded mainly during late spring.

`phenology_overlap_weights()` estimates the overlap between the day-of-year
distribution of a focal species and each comparison species using kernel-density
overlap. The overlap coefficient describes the common area of two estimated
distributions [@pastore2018; @pastore2019].

First, day of year is added to the bird records.

```{r phenology-data}
bird_dates <- aves_small
bird_dates$DoY <- as.POSIXlt(bird_dates$date)$yday + 1L

target_code <- bird_dates$code[
  match(target_species, bird_dates$species)
]

target_code
```

We then estimate phenological overlap between the focal species and every
species in the remaining target group.

```{r phenology-overlap}
phenology_weights <- phenology_overlap_weights(
  target = bird_dates,
  group = bird_dates[bird_dates$species != target_species, ],
  target_code = target_code,
  target_id = "code",
  group_id = "code",
  target_doy = "DoY",
  group_doy = "DoY",
  type = "2",
  min_n = 5
)

phenology_weights[
  order(phenology_weights$weight, decreasing = TRUE),
]
```

The returned `weight` values sum to one across comparison species with valid
overlap estimates. Here, each target-group observation receives the weight of
its species. Records are then summed within 100-m cells before KDE smoothing.

```{r phenology-weighted-points}
target_group_p <- merge(
  target_group,
  phenology_weights[, c("code", "weight")],
  by = "code",
  all.x = TRUE
)

target_group_p <- target_group_p[
  is.finite(target_group_p$weight) &
    !is.na(target_group_p$id100),
]

tg_weighted <- aggregate(
  list(weight = target_group_p$weight),
  by = list(id100 = target_group_p$id100),
  FUN = sum
)

idx <- match(tg_weighted$id100, points100$id)
keep <- !is.na(idx)

tg_points_weighted <- points100[idx[keep], ]
tg_points_weighted$weight <- tg_weighted$weight[keep]
```

```{r phenology-kde, fig.cap = "Phenologically weighted target-group sampling-bias surface."}
phenology_bias <- kde_surface(
  x = tg_points_weighted,
  weight_field = "weight",
  sigma = 3000,
  ref = example_grid,
  normalize = "meanG",
  mask = TRUE
)

terra::plot(
  phenology_bias,
  main = "Phenologically weighted target-group bias"
)
```

The weighted surface should be interpreted as a relative sampling-bias surface,
not as an estimate of species occurrence probability. Its purpose is to
describe where observations relevant to the focal species' seasonal sampling
process were more or less likely to have been collected.

# Seasonally and dielly weighted target group

For taxa whose detectability varies both seasonally and through the day,
seasonal overlap alone may be insufficient. `torus_overlap_many()` represents
season and time of day as two circular dimensions and estimates overlap on a
two-dimensional torus. Circular treatment is important because both annual and
daily cycles wrap around their endpoints; see @mardia2000 for directional
statistics and @ridout2009 for overlap estimation of diel activity patterns.


The bird dataset contains date-time observations.
Because this is an introductory vignette, the example uses the focal species
with the largest number of valid date-times and compares it with the 20 most
frequently recorded remaining species.

```{r torus-data}

target_species_torus <- target_species

reference_data <- aves_small[aves_small$species == target_species_torus & 
                               !is.na(aves_small$date_time),]

comparison_data <- aves_small[!is.na(aves_small$date_time),]

reference_data$obs_date <- as.Date(reference_data$date_time)
reference_data$obs_time <- hms::as_hms(reference_data$date_time)

comparison_data$obs_date <- as.Date(comparison_data$date_time)
comparison_data$obs_time <- hms::as_hms(comparison_data$date_time)
```

For speed, the example uses a coarser evaluation grid than would often be used
in a final analysis.

```{r torus-overlap}
torus_result <- torus_overlap_many(
  reference_data = reference_data,
  comparison_data = comparison_data,
  comparison_group_col = "species",
  reference_date_col = "obs_date",
  reference_time_col = "obs_time",
  comparison_date_col = "obs_date",
  comparison_time_col = "obs_time",
  reference_name = target_species_torus,
  season_bw_days = 14,
  daily_bw_hours = 1,
  date_resolution_days = 7,
  time_resolution_minutes = 60,
  min_n = 5,
  return_density = FALSE
)

torus_overlap <- torus_result$overlap
torus_overlap <- torus_overlap[
  order(torus_overlap$overlap, decreasing = TRUE),
]

head(torus_overlap, 10)
```

The overlap coefficient ranges from zero to one for normalized densities. A
large value indicates that a comparison species has a seasonal-diel observation
distribution similar to the focal species.

To use these coefficients as relative species weights in a target group, they
can be normalized across comparison species.

```{r torus-weights}
ok <- is.finite(torus_overlap$overlap)

torus_overlap$weight <- NA_real_
torus_overlap$weight[ok] <-
  torus_overlap$overlap[ok] /
  sum(torus_overlap$overlap[ok])

head(
  torus_overlap[
    order(torus_overlap$weight, decreasing = TRUE),
    c("comparison", "overlap", "weight")
  ],
  10
)
```

These weights can be joined back to the individual observations and passed to
`kde_surface()` in exactly the same way as the phenological weights above.

The overlap can also be visualised, when setting `return_density = TRUE`

```{r torus-plot}
torus_result <- torus_overlap_many(
  reference_data = reference_data,
  comparison_data = comparison_data,
  comparison_group_col = "species",
  reference_date_col = "obs_date",
  reference_time_col = "obs_time",
  comparison_date_col = "obs_date",
  comparison_time_col = "obs_time",
  reference_name = target_species_torus,
  season_bw_days = 14,
  daily_bw_hours = 1,
  date_resolution_days = 7,
  time_resolution_minutes = 60,
  min_n = 5,
  return_density = TRUE
)

ggplot(torus_result$density,aes(x = seasonal_position_days,
                          y = decimal_hour,
                          fill = comparison_density)) +
  geom_raster() +
  facet_wrap(~ comparison) +
  scale_y_continuous(breaks = seq(0, 24, by = 3)) +
  viridis::scale_fill_viridis() +
  labs(x = "Seasonal position (days from 1 January)",
       y = "Time of day",
       fill = "Density",
       title = "Species-specific registration activity") +
  theme_bw()

ggplot(torus_result$density,aes(x = seasonal_position_days,
                          y = decimal_hour,
                          fill = shared_density)) +
  geom_raster() +
  facet_wrap(~ comparison) +
  scale_y_continuous(breaks = seq(0, 24, by = 3)) +
  viridis::scale_fill_viridis() +
  labs(x = "Seasonal position (days from 1 January)",
       y = "Time of day",
       fill = "Density",
       title = "Registration activity overlap with STRALU") +
  theme_bw()


```


# Known inventory dates

Sometimes the dates of a survey or habitat inventory are known, while the
species itself was not necessarily recorded during that survey. In this case,
`doy_kde_weights()` can estimate how closely each inventory date corresponds to
the seasonal activity distribution of a target species.

The function estimates a wrapped day-of-year KDE from reference dates and
evaluates that density at another set of dates. The wrapped construction reduces
the artificial separation between observations near the end and beginning of a
year. Bandwidth controls the degree of smoothing, as in conventional KDE
[@silverman1986].

The example below uses one bryophyte species, which is considered old-growth 
forest specialist species, thus specifically searched during habitat inventories, as the
reference seasonal distribution and weights the dates in
`habitat_inventories`.

```{r known-dates}
bryophyta_species <- names(
  sort(table(bryophyta_one$species), decreasing = TRUE)
)[1]

reference_dates <- bryophyta_one$date[
  bryophyta_one$species == bryophyta_species &
    !is.na(bryophyta_one$date)
]

inventory_keep <- !is.na(habitat_inventories$date)

inventory_weights <- doy_kde_weights(
  A_dates = reference_dates,
  B_dates = habitat_inventories$date[inventory_keep],
  scale = "percentile_A",
  bw = "nrd0"
)

weighted_inventories <- habitat_inventories[inventory_keep, ]
weighted_inventories$phenology_weight <- inventory_weights$weights
```

With `scale = "percentile_A"`, weights describe how high the estimated
reference density is relative to density values across the seasonal cycle.
They are relative phenological correspondence scores rather than occurrence
probabilities.

```{r inventory-summary}
aggregate(
  phenology_weight ~ habitat_type,
  data = weighted_inventories,
  FUN = mean
)
```

The estimated seasonal reference curve can also be inspected directly.

```{r inventory-density, fig.cap = "Estimated seasonal activity curve used for weighting known inventory dates."}
plot(
  inventory_weights$density$x,
  inventory_weights$density$y,
  type = "l",
  xlab = "Day of year",
  ylab = "Estimated reference density"
)
```



# Selecting an independent testing set

Random train-test splitting can give overly optimistic estimates of predictive
performance when nearby observations are spatially dependent. Spatially
structured resampling is therefore generally preferable when the intended
prediction involves geographic transfer [@roberts2017].

`sdmhelpers` does not define the spatial blocks themselves. Instead,
`select_folds_by_quota()` and `select_joint_folds()` select complete groups from
fold identifiers created elsewhere. Independence therefore comes from the
design of the folds, not from the selection function.

The package contains example presence records, background records, and
independent spatial polygons.

```{r spatial-folds}
pres_sf <- sf::st_as_sf(
  example_presences,
  coords = c("x", "y"),
  crs = 3059
)

bg_sf <- sf::st_as_sf(
  example_background,
  coords = c("x", "y"),
  crs = 3059
)

block_id <- setdiff(
  names(example_independent_blocks),
  attr(example_independent_blocks, "sf_column")
)[1]

pres_joined <- sf::st_join(
  pres_sf,
  example_independent_blocks[block_id],
  join = sf::st_within,
  left = TRUE
)

bg_joined <- sf::st_join(
  bg_sf,
  example_independent_blocks[block_id],
  join = sf::st_within,
  left = TRUE
)

pres_folds <- pres_joined[[block_id]]
bg_folds <- bg_joined[[block_id]]
```

When both presences and background should be withheld from the same spatial
blocks, `select_joint_folds()` is convenient.

```{r joint-fold-selection}
independent <- select_joint_folds(
  pres_folds = pres_folds,
  bg_folds = bg_folds,
  target_prop = 0.25,
  pres_bounds = c(0.20, 0.30),
  bg_bounds = c(0.20, 0.30),
  max_tries = 3000,
  bg_match_weight = 1,
  seed = 1,
  print_report = FALSE
)

independent$selected_fold_ids
independent$summary[c(
  "n_pres_selected",
  "n_bg_selected",
  "prop_pres_selected",
  "prop_bg_selected"
)]
```

The returned logical vectors preserve the row order of the input data and can
be used directly for subsetting.

```{r split-independent}
pres_test <- example_presences[independent$pres_selected_idx, ]
pres_train <- example_presences[!independent$pres_selected_idx, ]

bg_test <- example_background[independent$bg_selected_idx, ]
bg_train <- example_background[!independent$bg_selected_idx, ]

c(
  pres_train = nrow(pres_train),
  pres_test = nrow(pres_test),
  bg_train = nrow(bg_train),
  bg_test = nrow(bg_test)
)
```

If only one type of observation needs to be selected, the simpler
`select_folds_by_quota()` can be used instead.

```{r one-set-fold-selection}
presence_only_selection <- select_folds_by_quota(
  folds = pres_folds,
  target_prop = 0.25,
  bounds = c(0.20, 0.30),
  seed = 1,
  print_report = FALSE
)

presence_only_selection$summary$prop_selected
```

# Screening predictor variance

Predictors with zero or extremely small variance in the data available for
model fitting can create numerical problems and cannot meaningfully contribute
to estimating an environmental response. `screen_egv_variance()` checks
predictor variance across several SDM-relevant subsets rather than only across
the complete dataset.

The example environmental raster contains five ecogeographical variables.

```{r egv}
env <- terra::rast(egv_file)
env
```

For a compact vignette example, a subset of the training background is used to
construct an `SDMtune::SWD` object.

```{r swd,eval=TRUE}
set.seed(42)

bg_screen <- bg_train[sample.int(nrow(bg_train), 5000), ]

train_swd <- SDMtune::prepareSWD(
  species = "example_species",
  p = pres_train,
  a = bg_screen,
  env = env
)
```

The environmental-domain check is added by supplying `env`.

```{r variance-screen, eval=TRUE}
variance_screen <- screen_egv_variance(
  train = train_swd,
  env = env,
  env_sample_n = 5000,
  seed = 1,
  sd_tol = 0,
  verbose = FALSE
)

variance_screen$keep
variance_screen$remove
```

A predictor is retained only if it passes every requested check. The
`diagnostics` component provides the standard deviation and failure reason for
every evaluated predictor/subset combination.

```{r variance-diagnostics, eval=TRUE}
head(variance_screen$diagnostics)
```


# Fitting a small maxnet model

The package is designed to support, rather than replace, SDM software. The
following compact example fits a `maxnet` model after the independent split
created above. Only predictors retained by the variance screen are used.

First extract environmental values for the training and independent-testing
locations.

```{r extract-env}
training_set <- SDMtune::prepareSWD(
  species = "example_species",
  p = pres_train,
  a = bg_train,
  env = env
)
training_set=SDMtune::addSamplesToBg(training_set)

testing_set <- SDMtune::prepareSWD(
  species = "example_species",
  p = pres_test,
  a = bg_test,
  env = env
)
testing_set=SDMtune::addSamplesToBg(testing_set)

```


Fit the model using presences (`1`) and background (`0`).

```{r fit-maxnet}
maxnet_model <- maxnet::maxnet(
  p = training_set@pa,
  data = training_set@data
)
```


# Evaluating models with TSS

The true skill statistic is

\[
\mathrm{TSS} = \mathrm{sensitivity} + \mathrm{specificity} - 1,
\]

and was proposed for SDM evaluation as an alternative to measures such as
Cohen's kappa [@allouche2006].

`max_tss()` evaluates all unique finite prediction thresholds and returns the
threshold that maximizes TSS together with sensitivity and specificity.

For presence-background models, background locations are not confirmed
absences. Consequently, the value reported as specificity is operationally the
proportion of background records classified as negative, and the resulting TSS
should be interpreted as presence-background discrimination rather than
classification accuracy against known true absences.

```{r independent-predictions}
p_pos <- predict(
  maxnet_model,
  newdata = testing_set@data[testing_set@pa == 1, ],
  type = "cloglog"
)

p_neg <- predict(
  maxnet_model,
  newdata = testing_set@data[testing_set@pa == 0, ],
  type = "cloglog"
)

tss_result <- max_tss(
  p_pos = p_pos,
  p_neg = p_neg
)

tss_result
```

The selected threshold can then be used to convert continuous predictions into
binary classifications if a binary map or confusion matrix is required.

```{r apply-threshold}
threshold <- unname(tss_result["threshold"])

presence_class <- p_pos >= threshold
background_class <- p_neg >= threshold

c(
  sensitivity = mean(presence_class),
  background_specificity = mean(!background_class)
)
```

# Cross-validation TSS with ENMeval

The independent testing set created above is reserved for evaluating the final
model and should not be used during model tuning. Alternative model settings
can instead be compared by cross-validation using only the training data.
Spatially structured cross-validation is particularly useful when observations
are spatially autocorrelated because it reduces spatial dependence between
calibration and validation data and provides a more stringent assessment of
spatial transferability than random partitioning [@roberts2017].

ENMeval provides several approaches for partitioning occurrence and background
records and for evaluating alternative Maxent and `maxnet` model settings
[@muscarella2014; @kass2021]. In this example, the training records are divided
into four geographically structured groups using `ENMeval::get.block()`.

```{r}
block_folds <- ENMeval::get.block(occ = training_set@coords[training_set@pa == 1, ], 
                                  bg = training_set@coords[training_set@pa == 0, ])
```

The block method divides occurrence localities into four spatial groups and
assigns background records to the corresponding geographic partitions. Because
the example coordinates are in the projected LKS-92 / Latvia TM coordinate
reference system, the partitions are defined using projected X and Y
coordinates rather than geographic longitude and latitude.

The resulting fold identifiers are supplied to `ENMevaluate()` as user-defined
partitions. This makes it possible to use the same spatial groups for both
presence and background observations while retaining explicit control over the
cross-validation design.

For this example, three feature-class combinations and three regularization
multipliers are evaluated, producing nine candidate model settings. Linear
(`L`), quadratic (`Q`), and combined linear-quadratic (`LQ`) feature classes
control the shapes of fitted environmental responses, whereas the
regularization multiplier controls the strength of penalization and therefore
model complexity.

```{r fit-maxnet-ENMeval}
presences=cbind(training_set@coords[training_set@pa == 1, ],
                 training_set@data[training_set@pa == 1, ])
background=cbind(training_set@coords[training_set@pa == 0, ],
           training_set@data[training_set@pa == 0, ])
pirmais <- ENMeval::ENMevaluate(occs = presences, 
                       bg = background,
                       algorithm = 'maxnet', 
                       partitions = 'user', 
                       user.grp=list(occs.grp=block_folds$occs.grp,
                                     bg.grp=block_folds$bg.grp),
                       tune.args = list(fc = c('L','Q','LQ'), 
                                        rm = c(0.5,1,2)),
                       other.settings=list(validation.bg='partition'),
                       user.eval = tss_user_eval)
```

The argument `user.eval = tss_user_eval` adds TSS as a custom evaluation metric.
For every cross-validation fold, `ENMeval` supplies predictions for training 
occurrences, validation occurrences, training background records and validation 
background records to `tss_user_eval()`, which then applies `max_tss()` to these validation
predictions and returns the maximum TSS and its associated classification
threshold.

When

```{r,eval=FALSE}

other.settings = list(validation.bg = "partition")
```


is used, validation occurrences are compared only with background records from
the corresponding validation partition. This means that both presences and
background records are spatially withheld during evaluation. If
`validation.bg = "full"` is used instead, validation occurrences are compared
with the combined training and validation background predictions.

For presence-background models, background locations are not confirmed
absences. Consequently, the reported specificity is the proportion of
background predictions falling below the selected threshold rather than
specificity calculated from known true absences. The resulting TSS should
therefore be interpreted as a measure of discrimination between presences and
background rather than conventional classification accuracy.

`tss_user_eval()` selects the threshold maximizing TSS separately for the
training and validation predictions. Consequently, both `tss.train` and
`tss.val` are optimized on the same data on which they are evaluated. In
particular, `tss.val` should therefore be regarded as a cross-validation metric
for comparing candidate models rather than as a fully independent estimate of
threshold-based predictive performance.

The model-level results returned by `ENMeval` summarize the fold-level
statistics and can be inspected with:


```{r}
ENMeval::eval.results(pirmais)
```



Fold-specific results can be examined with:


```{r}

ENMeval::eval.results.partitions(pirmais)
```


Cross-validation TSS and independent-test TSS serve different purposes in this
workflow. Cross-validation is used to compare candidate feature classes,
regularization multipliers or other model settings using only the training
data. Once the preferred settings have been selected and the final model has
been fitted, the completely withheld independent testing set can be used for a
separate assessment of predictive performance.

For a strictly independent threshold-based assessment, the classification
threshold should be selected from training or cross-validation data and then
applied unchanged to the independent testing data, rather than re-optimizing
the threshold on the independent test set.


# References
